Causal inference and co-evolutionary contingencies using dynamic phylogenetic models

### 2022-07-13 manuscript.qmd
round off phyo confounding section

First, what do we mean by causal? Following Pearl (some citations), we say that $X$ is a cause of $Y$ when $P(Y|do(X = x)) \neq P(Y|do(X = \textrm{not x}))$. The do($\dot$) operator represents an *intervention* in which the value of some variable is fixed (for example, through assignment in a randomized experiment). This is an inclusive definition, where if the probability distribution $P(Y)$ is sensitive to the value of $X$ that is assigned, then $X$ has a causal effect on $Y$. Note here that $X$ refers to a random variable while $x$ refers to some realized unit value of that variable. In any observational study, we cannot directly access $P(Y|do(X = x))$. Instead, we have just have the condition probability of $Y$ given $X$, $P(Y|X)$. Pearl and colleagues' "do calculus" is a set of rules that tells us under what conditions $P(Y|do(X)) = P(Y|X)$, or when we can make them equivalent by adjustment for other variables $P(Y|do(X),Z) = P(Y|X,Z)$. An *estimand* is a unit specific estimate such as "what would we expect species $i$ trait value $y_i$ to be if $x_i$ took on a given value?". Under a particular set of causal and parameteric assumptions, a regression coefficient $b$ can be interpeted as the estimand "what is the difference in the expected (mean) value of $Y$ if $x = 1$ vs $x = 0~$?", or $E[Y|x=1] - E[Y|x=0]$.

@@

We previously stated that a cross section of current species trait values can provide an unbiased estimate of the change in trait $Y$ in response to changes in $X$. This fact is well known in the comparative literature [@pagel1993seeking; @rohle2006comment], and consistent with the rules of do-calculus--given the strict assumptions of the model outined before. However, if traits were correlated in the past but no longer exhibit any association, then evolutionary history is a confounder. For example, in Figure 4B, $U_0$ creates a "backdoor path" that connects all pairs of $X_t$ and $Y_t$.

@@

HOW CAN WE TELL IF ITS PHYLOGENETIC CONFOUNDING VS A CAUSAL RELATIONSHIP? IN THE FORMER, THE COVARIANCE BETWEEN X AND Y WILL DECREASE OVER TIME. IN THE LATTER IT IS CONSTANT/STABLE.

Additionally, even if there was zero structural phylogenetic confounding, it would still be a good idea in any finite sample to adjust for phylogeny because, as a source of variation in $Y$, phlyogeny is a "good control" in that it helps us isolate variation in $Y$ that is due to possible causal effects of $X$. Marginalizing over phylogenetic position can also give us improved estimates of the average effects when clades are unequally sampled.

This model assumes that trait co-evolution is *ergodic*, which means that the time average of the co-evolutionary process is equal to the ensemble average. In other words, a cross-section contains the same information as a time series. Remarkably, this means that even a simple species regression (i.e., regression of $Y$ on $X$, with no phylogenetic information considered) provides an unbiased estimate of $b$.

An OU model where the response is undergoing adaptive evolution and the predictor is drifting also works. But if the predictor is under selection, things get messed up.

## Causality on Phylogenetic Trees

![Figure 1](figures/phylo_DAG.png)

@@

Brownian motion is the simplest and most widely used model, and often interpreted as "drift" across the phylogenetic tree. Other popular choices include Ornstein-Uhlenbeck, where change is due to both random drift and "stabilizing selection" around some optimal trait value, as well as Early Burst models of adaptive radiation where the rate of change ($\sigma$) decreases over time. In general, $\boldsymbol{V}$ is constructed by inputting a matrix of phylogenetic distances to some covariance function $C(x,x')$ such that $V_{[j,k]} = C(j,k)$.

@@

#### Multivariate PGLMMs

@@

#### Phylogenetic Path Analysis & Causal Inference

@@

## "Dynamic" PCMs

>>>>

First, what do we mean by causal? Following Pearl (some citations), we say that $X$ is a cause of $Y$ when $P(Y|do(X = x)) \neq P(Y|do(X = \textrm{not x}))$. The do($\cdot$) operator represents an *intervention* in which the value of some variable is fixed (for example, through assignment in a randomized experiment). This is an inclusive definition, where if the probability distribution $P(Y)$ is sensitive to the value of $X$ that is assigned, then $X$ has a causal effect on $Y$. Note here that $X$ refers to a random variable while $x$ refers to some realized unit value of that variable. In any observational study, we cannot directly access $P(Y|do(X = x))$. Instead, we have just have the condition probability of $Y$ given $X$, $P(Y|X)$. Pearl and colleagues' "do calculus" is a set of rules that tells us under what conditions $P(Y|do(X)) = P(Y|X)$, or when we can make them equivalent by adjustment for other variables $P(Y|do(X),Z) = P(Y|X,Z)$. An *estimand* is a unit specific estimate such as "what would we expect species $i$ trait value $y_i$ to be if $x_i$ took on a given value?". Under a particular set of causal and parameteric assumptions, a regression coefficient $b$ can be interpeted as the estimand "what is the difference in the expected (mean) value of $Y$ if $x = 1$ vs $x = 0~$?", or $\mathbb{E}[Y|x=1] - \mathbb{E}[Y|x=0]$.

@@

We previously stated that a cross section of current species trait values can provide an unbiased estimate of the change in trait $Y$ in response to changes in $X$. This fact is well known in the comparative literature [@pagel1993seeking; @rohle2006comment], and consistent with the rules of do-calculus--given the strict assumptions of the model outlined before. However, if traits were correlated in the past but no longer exhibit any association, then evolutionary history acts as a confounder. For example, in Figure 4B, $U_0$ creates a "backdoor path" that connects all pairs of $X_t$ and $Y_t$. This means that $P(Y|do(X)) \neq P(Y|X)$, unless we can somehow adjust for $U_0$.

The confounder denoted $U$ is unobserved, meaning that we can't statistically adjust for it. However, note that for any time $t$, $X_t$ and $Y_t$ are independent of $U_0$ conditional on $\{ X_{t-1},X_{t-1}$. In other words, the present is conditionally independent of $U_0$ (or "d-seperated") given the past. But in most comparative studies, we have not observed the past, only the present. This is where phylogeny can help us, because more closely related species or populations have more similar confounding factors (e.g., shared genetic architecture, environmental similarity), on average, than distantly related ones. Assuming that we have an appropriate model of evolution, then the patristic distance between species (i.e., time since divergence from a common ancestor) allows us to probabalistically reconstruct unobserved past values of $X$ and $Y$, closing the backdoor path $X \leftarrow U \rightarrow Y$ (cite Rethinking 2nd edition). It should be noted that such a scenario is sometimes called "phylogenetic confounding," but...[DOESN"T WORK IF U IS STILL PRESENT AT THE TIPS. HOW CAN WE TELL IF ITS PHYLOGENETIC CONFOUNDING VS A CAUSAL RELATIONSHIP? IN THE FORMER, THE COVARIANCE BETWEEN X AND Y WILL DECREASE OVER TIME. IN THE LATTER IT IS CONSTANT/STABLE].

@@

Additionally, even if there was zero structural phylogenetic confounding (Figure 1A), it would still be a good idea in any finite sample to adjust for phylogeny because, as a source of variation in $Y$, phlyogeny is a "good control" in that it helps us isolate variation in $Y$ that is due to possible causal effects of $X$. Marginalizing over phylogenetic position can also give us improved estimates of the average effects when clades are unequally sampled.

Now that we have a sense for the assumed causal model underlying most phylogenetic comparative analyses, and motivation for why we statistically adjust for phylogeny, let's review some popular methods for estimating the evolutionary regression coefficient and introduce some variations on the causal model in Figure 1.

@@

Brownian motion is the simplest and most widely used model, and often interpreted as "drift" across the phylogenetic tree. Other popular choices include Ornstein-Uhlenbeck, where change is due to both random drift and "stabilizing selection" around some optimal trait value, as well as Early Burst models of adaptive radiation where the rate of change ($\sigma$) decreases over time. NOTE ON OU COVARIANCE: An OU model where the response is undergoing adaptive evolution and the predictor is drifting also works. But if the predictor is under selection, things get messed up. In general, $\boldsymbol{V}$ is constructed by inputting a matrix of phylogenetic distances to some covariance function $C(x,x')$ such that $V_{[j,k]} = C(j,k)$.

@@

### Mutli-response PGLMMs: contrasting macro and microevolution

@@

### Misc. Notes

OU BUTLER AND KING ALGO--PIECEWISE CONSTANT SELECTION REGIME. I THINK THIS MEANS THAT Y RESPONDS TO WHATEVER TRIAT VALUE X WAS AT THE START OF THE SEGMENT--DOESN'T RESPOND TO CHANGES IN X WITHIN THE SEG.

![Figure 1](figures/phylo_DAG.png)
### Phylogenetic Path Analysis

@@

## Dynamic state-space models of co-evolution
